---
title: "Moving Democrats for the Senate"
author: "Daniella Raz"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)

# from David Shor's calculations, including std. dev.
shor_data_trimmed = read.csv("shor_data_trimmed.csv")
std_dev = 0.0362
shor_data_trimmed$baseline_winning = pnorm(shor_data_trimmed$expected_share, mean = 0.5, sd = std_dev)
```

input:

  1. dem: the number of votes received by Joe Biden in 2020

  2. repub: the number of votes received by Donald Trump in 2020

  3. old_probability_win: the probability of the Democrat winning the state in a neutral environment, given their two-party vote share 

  4. expected_prop: the expected proportion of the two-party vote won by the Democrat in a neutral environment

  5. std_dev: the standard deviation of two-party vote share 

  6. votes: the number of votes added in this iteration
  
output:

  1. updated_dem: the number of votes received by Joe Biden in 2020, adjusted for the votes we moved (same as dem for 48/50 states)
  
  2. updated_probability_win: old_probability_win updated with the shifted vote total (same as old_probability_win for 48/50 states)
  
  3. updated_expected_prop: expected_prop updated with shifted vote total (same as expected_prop for 48/50 states)
  
  4. new_seats: the expected number of senate seats under this allocation
```{r}
demo_vote_adder = function(dem, repub, old_probability_win, expected_prop, std_dev, votes){
  
  # Calculate the new expected proportion of the democrat vote in a neutral environment if each state 
  # added or subtracted "votes" 
  dem_prop = (expected_prop * (dem + repub) + votes) / (dem + repub + votes)
  dem_prop_subtract = (expected_prop * (dem + repub) - votes) / (dem + repub - votes)
  
  # Calculate the updated probability of winning in a neutral environment if each state added or subtracted "votes"
  new_probability_win = pnorm(dem_prop, mean = 0.5, sd = std_dev)
  new_probability_win_subtract = pnorm(dem_prop_subtract, mean = 0.5, sd = std_dev)
  
  # Calculate how much that probability changed in each state
  change_in_prob = new_probability_win - old_probability_win
  change_in_prob_subtract = new_probability_win_subtract - old_probability_win
  
  # Finding which state saw the biggest increase with "votes" added and the smallest decrease
  # with "votes" subtracted. 
  # Note, I prevent taking additional votes from a state with 0% democrats. 
  max_change = max(change_in_prob)
  max_change_subtract = max(change_in_prob_subtract - as.numeric(dem_prop_subtract < 0)*100)
  
  # Create variables that show the change in Dem win probability after adding votes to 1 state and subtracting
  # votes from another state.
  state_add_pct = as.numeric(change_in_prob == max_change) * max_change
  state_subtract_pct = as.numeric(change_in_prob_subtract == max_change_subtract) * max_change_subtract
  
  # Update win probabilities and expected proportion of two-party vote won by Democrats in a neutral environment after
  # shifting votes from one state to another.
  updated_probability_win = old_probability_win + state_add_pct + state_subtract_pct
  updated_dem = dem + as.numeric(change_in_prob == max_change) * votes - as.numeric(change_in_prob_subtract == max_change_subtract) * votes
  updated_expected_prop = as.numeric(change_in_prob == max_change) * dem_prop + 
    as.numeric(change_in_prob_subtract == max_change_subtract) * dem_prop_subtract + 
    as.numeric(change_in_prob != max_change & change_in_prob_subtract != max_change_subtract) * expected_prop
  
  # Calculate the expected number of seats won by Dems in a neutral environment given the updated vote allocations.
  # This is equal to the sum of probability of winning each state, multiplied by 2. 
  new_seats = sum(updated_probability_win)*2
  return(list(updated_dem, updated_probability_win, updated_expected_prop, new_seats))
}
```


I set it to iterate through 200,000 votes at a time, at first. We use the base values from David
Shor for the first iteration. In a neutral environment, Democrats should expect to
win 39.62 Senate seats before re-allocating votes. 
```{r}
# Expected Senate seats won
sum(shor_data_trimmed$baseline_winning)*2

vote_increments = 200000

votes_added = demo_vote_adder(shor_data_trimmed$biden, 
                              shor_data_trimmed$trump, 
                              shor_data_trimmed$baseline_winning,
                              shor_data_trimmed$expected_share,
                              std_dev,
                              vote_increments)

updated_dem = votes_added[[1]]
updated_probability_win = votes_added[[2]]
updated_expected_prop = votes_added[[3]]
new_seats = votes_added[[4]]
vote_counter = vote_increments
```

Going through 4 iterations of 200,000 vote chunks. 
Then continue at 5000 vote chunks until Democrats are expected to win 50 seats. 
Note that across the past 6 presidential elections (2000-2020), Democrats have
won an average of 51.4% of the two-party vote share. 
To allocate votes with that threshold inmind, change 50 to 51.4 in line 132.
```{r}
counter = 1
while(counter < 5){
  votes_added = demo_vote_adder(updated_dem, 
                                shor_data_trimmed$trump,
                                updated_probability_win, 
                                updated_expected_prop,
                                std_dev, 
                                vote_increments)
  
  updated_dem = votes_added[[1]]
  updated_probability_win = votes_added[[2]]
  updated_expected_prop = votes_added[[3]]
  new_seats = votes_added[[4]]
  vote_counter = vote_counter + vote_increments
  
  counter = counter + 1
}

vote_increments = 5000
while(new_seats < 50){
  votes_added = demo_vote_adder(updated_dem, 
                                shor_data_trimmed$trump,
                                updated_probability_win, 
                                updated_expected_prop,
                                std_dev, 
                                vote_increments)
  
  updated_dem = votes_added[[1]]
  updated_probability_win = votes_added[[2]]
  updated_expected_prop = votes_added[[3]]
  new_seats = votes_added[[4]]
  vote_counter = vote_counter + vote_increments
}
```


```{r, echo = FALSE}
# Compile the results.
results_df = data.frame(state = shor_data_trimmed$state, 
                        votes_added = updated_dem - shor_data_trimmed$biden,
                        original_expected_prop = round(shor_data_trimmed$expected_share, 3),
                        updated_expected_prop = round(updated_expected_prop,3), 
                        original_probability_win = round(shor_data_trimmed$baseline_winning, 4),
                        updated_probability_win = round(updated_probability_win, 4))

```

Results: 

```{r}
total_num_votes_moved = sum(abs(results_df$votes_added)) / 2
total_num_votes_moved
new_seats
results_df

```

Chart 1: Expected Democrat seats in different national environments

```{r}
change_from_50 = seq(-.1, .1, by = 0.0025)
national_dem_vote = 0.5 + change_from_50

expected_seats = rep(0, length(change_from_50))
for(i in 1:length(change_from_50)){
  updated_dem_prop = shor_data_trimmed$expected_share + change_from_50[i]
  updated_dem_win_pct = pnorm(updated_dem_prop, mean = 0.5, sd = std_dev)
  expected_seats[i] = sum(updated_dem_win_pct)*2
}

seat_environment = data.frame(expected_seats, national_dem_vote)
write.csv(seat_environment, "seat_environment.csv", row.names = FALSE)
```

Chart 2: Map showing Democratic voters transferred out of CA to recipient states

```{r}
# adjusting for removal of all votes from CA instead
# transferring votes removed from other states to CA (total 1.1 million votes)
results_df %>% mutate(updated_expected_prop = case_when(votes_added < 0 ~ original_expected_prop,
                                                        TRUE ~ updated_expected_prop),
                      updated_probability_win = case_when(votes_added < 0 ~ original_probability_win,
                                                          TRUE ~ updated_probability_win),
                      votes_added = case_when(votes_added < 0 ~ 0,
                                              state == "CA" ~ -1100000,
                                              TRUE ~ votes_added)) -> results_df

ca_prop_subtract = (shor_data_trimmed$expected_share[40] * (shor_data_trimmed$biden[40] + shor_data_trimmed$trump[40]) - 1100000) / 
  (shor_data_trimmed$biden[40] + shor_data_trimmed$trump[40] - 1100000)

# calculating for CA the updated probability of winning in a neutral environment if each state added or subtracted "votes"
ca_probability_win = pnorm(ca_prop_subtract, mean = 0.5, sd = std_dev)
 
results_df$updated_expected_prop[40] = ca_prop_subtract
results_df$updated_probability_win[40] = ca_probability_win
  
sum(results_df$updated_probability_win) * 2

write.csv(results_df, "data_for_map.csv", row.names = FALSE)
```

Chart 3: Stacked bar chart showing state-level

```{r}
stacked_chart_data = data.frame(state = shor_data_trimmed$state,
                                neutral_dem_votes = shor_data_trimmed$expected_share * 
                                  (shor_data_trimmed$biden + shor_data_trimmed$trump),
                                neutral_repub_votes = (1 - shor_data_trimmed$expected_share) *
                                  (shor_data_trimmed$biden + shor_data_trimmed$trump),
                                votes_moved = results_df$votes_added,
                                probability_dem_win = results_df$updated_probability_win)

stacked_chart_data %>% mutate(rating = case_when(updated_probability_win <= 0.01 ~ "safe republican",
                                                 updated_probability_win > 0.01 & updated_probability_win <= 0.1 ~ "solid republican",
                                                 updated_probability_win > 0.1 & updated_probability_win <= 0.25 ~ "likely republican",
                                                 updated_probability_win > 0.25 & updated_probability_win <= 0.4 ~ "lean republican",
                                                 updated_probability_win > 0.4 & updated_probability_win <= 0.6 ~ "toss up",
                                                 updated_probability_win > 0.6 & updated_probability_win <= 0.75 ~ "lean democrat",
                                                 updated_probability_win > 0.75 & updated_probability_win <= 0.9 ~ "likely democrat",
                                                 updated_probability_win > 0.9 & updated_probability_win <= 0.99 ~ "solid democrat",
                                                 updated_probability_win > 0.99 ~ "safe democrat")) -> stacked_chart_data

write.csv(stacked_chart_data, "stacked_chart_data.csv", row.names = FALSE)

```


